---
title: "Foundations of R programming II"
author: "Shamit Soneji"
date: "12/10/2018"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Shit gets real now.....
In the first R workshop you were shown the basics of the R language, subsetting vector and matricies, writing functions and basic plotting. Today we're going to roll all that together and start looking at how to wrap all these things up into an R package that will use S4 classes that you will make available on github.

This will be a fairly intense workshop, but if you can crack it will boost your knowledge and confidence.

The first thing we'll do is revisit `lists` from the first workshop. Here is the code we used to make one:

```{r}
alpha <- LETTERS[1:8]
mat <- matrix(rnorm(40),nrow=8)
listex1  <- list(char=alpha,nums=mat)
listex1
```

You will rememver that elements of a list can be accessed using the `$` character, so'
```{r}
listex1$mat
```
and subset in the usual way:
```{r}
listex1$mat[2:3,] #prints just rows 2 and 3
```